Model for DES Type Ia Supernova Cosmology Analysis

### 2015-07-20 doc/model.tex
Incorporate Danny's comments

\documentclass[preprint]{aastex}
\usepackage{amsmath,amssymb}
\usepackage{mathrsfs}
\usepackage{graphicx}
\usepackage{natbib}
\bibliographystyle{apj}

@@

\title{Type Ia Supernova Model}
\author{Alex Kim}

\section{Model}
We infer a Hubble diagram based on measurements of discovered transients
classified as Type~Ia supernovae (SNe~Ia).  The probability distribution
function (pdf) of the Hubble diagram is denoted as
\begin{equation}
P(\vec{\mu},\vec{z} |  \vec{\hat{ADU}}, \vec{\hat{T}}_S,\vec{\hat{z}}_S,
\vec{\hat{\theta}}_S,\hat{\text{Gals.}}, \vec{\hat{D}}=1, \vec{\hat{T}}=\text{SN~Ia},
\vec{\hat{\text{RA}}}, \vec{\hat{\text{Dec}}}),
\label{hd:eqn}
\end{equation}
where the inferred quantities are the set of distance moduli ($\vec{\mu}$)
and redshifts ($\vec{z}$), and 
the measurements include:  the transient photometry $\hat{ADU}$;
spectroscopically derived transient
typing, redshift and parameters $\hat{T}_S$, $\hat{z}_S$, $\hat{\theta}_S$;
and the galaxy catalog $\hat{\text{Gals.}}$
The conditions $\vec{\hat{D}}=1$, $\vec{\hat{T}}=\text{SN~Ia}$ indicate a sample selection
of transients
discovered and classified as SNe~Ia. The observed coordinates of the transients
are $\vec{\hat{\text{RA}}}$, $\vec{\hat{\text{Dec}}}$.
We use the convention of using carets $\hat{}$
to  label measurements.

The pdf in Equation~\ref{hd:eqn} is proportional to
\begin{equation}
P(\vec{\mu},\vec{z},  \vec{\hat{D}}=1, \vec{\hat{T}}=\text{SN~Ia},
\vec{\hat{\text{RA}}}, \vec{\hat{\text{Dec}}} |  \vec{\hat{ADU}}, \vec{\hat{T}}_S,\vec{\hat{z}}_S,
\vec{\hat{\theta}}_S,\hat{\text{Gals.}})
\label{hd2:eqn}
\end{equation}
and we model this scenario.  A sketch of such a model is depicted in the Probabilistic Graphical Model
(PGM)
shown in Figure~\ref{pgm:fig}.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=7in]{/Users/akim/project/abc/results/hdpgm.eps} 
   \caption{Probabilistic Graphical Model for the SN~Ia analysis.  
   The forward modeling
   flow goes from left to right, starting with the description of the universe, observatory,
   and data.
   A galaxy catalog $\hat{\mathit{Gal.}}$ informs the model for potential hosts $G$.  A transient
   at coordinates  $\text{RA/Dec}$  has a host $g_i$ drawn from $G$.
   The galaxy redshift redshift $z$ and parameters $\theta_{gi}$ are fixed based on the host.
   A distance modulus model with parameters $\theta_\mu$ fix the distance modulus $\mu_i$.
   The transient type $T$ depends on the host-galaxy.   The transient
   parameters $\theta_T^X$, $\theta_{Ti}^X$ (and perhaps the host galaxy) fix the luminosity $L$.       The 
   incoming photon flux $n$, $n_g$  are then fixed
   with redshift and distance modulus.
   The instrumental transmission function $\phi$ gives the expected
   counts $\mathit{ADU}$, $\mathit{ADU}_g$.  The calibration measurements
   are $\hat{Z}$.  The measured galaxy catalog is $\hat{\mathit{Gals.}}$  Only detected
   supernovae $\hat{D}$ have subsequent light curve ($\hat{ADU}$), spectral
   $\hat{X}_S$, and associated host galaxy $\hat{X}_H$ measurements.  A supernova
   classifier takes measurements and infers a type $\hat{T}$, which is used for sample selection.
       Nodes labeled by
   \textasciicircum are fixed, the others are probilistic.
   \label{pgm:fig}}
\end{figure}

The supernova survey provides the transients that are to be analyzed providing
the following information:
\begin{itemize}
\item $\hat{\text{RA}}$, $\hat{\text{Dec}}$.  The observed
coordinates.  For our purposes the observed, rather than true, coordinates are sufficient.
\item $\hat{D}=1$.  The transient is discovered.
\item $\hat{T}=\text{SN~Ia}$. The transient is classified as SN~Ia.
\item $\hat{\text{Gals}}$ is a galaxy catalog used to make host-galaxy associations.
\end{itemize}
In the PGM, these parameters are labeled as fixed ($\bullet$) data (\textasciicircum).

The model describes the universe and its contents, specifically transients
that are discovered as typed as SN~Ia and their host galaxies.
\begin{itemize}
\item $G$ is a shorthand for the model of all galaxies that can host a transient.  It is
the union of the objects in the observed galaxy catalog $\hat{\text{Gals}}$
plus unobserved galaxies that have not
met the detection threshold, presumably due to their low surface brightness.
\item $P(g_i | G, \hat{\text{RA}}/\hat{\text{Dec}})$. The host galaxy $g_i$
is drawn from the galaxy catalog based on coordinates.
This pdf is
the deliverable we want from Ravi.
\item $z_i$, $\theta_{gi}$ are the galaxy redshift and parameters.
For observed galaxies the parameter values are based on measurements e.g.\ $\theta_{gi} \sim \mathcal{N}(\hat{\theta}_{gi},\sigma^2_{\hat{\theta}_{gi}})$, otherwise the distribution
of $\theta_{gi}$ for ``hostless'' galaxies is described by the model.
Redshifts may be photometric or spectroscopic.
\item $\theta_\mu$ are the parameters that specify that distance modulus.
\item $\mu_i=\mu(z_i; \theta_\mu)$.  The distance modulus is a function of redshift,
we don't impose a specific form for  $\mu$ at the moment,  it could
be a $\Lambda$CDM prediction with cosmological parameters, a spline with knot values
as parameters, or as extreme as independent values for each $z$.
The redshift of the galaxy is used in the Hubble diagram.
\item $P(T_i | \theta_{gi})$.  The transient type $T_i$, SN~Ia or non-Ia, depends
on the host properties.  We could make the relative rates be dependent
on specific star formation rate and morphology.
\item $P(\theta_T^{Ia}, \theta_{Ti}^{Ia})$.  The model parameters for SNe~Ia, including
parameters that describe distributions.
Different SN~Ia models could be considered: the treatment of SALT2 parameters
has been presented in \citet{2011MNRAS.418.2308M}.
The SALT2 SN~Ia
model has individual $c$ and $x_1$ parameters that would constitute $\theta_{Ti}^{Ia}$,
and global $\alpha$, $\beta$, and intrinsic dispersion  parameters  common to all supernovae
 and  underlying  $c$ and $x_1$ distributions that would constitute $\theta_T^{Ia}$.
\item $P(\theta_T^{non-Ia}, \theta_{Ti}^{non-Ia})$.  Non-Ia transients
that can be mistyped as Ia are described by a parameterized model.
\item $L_i(t,\lambda; T_i, \theta_T^{Ia}, \theta_{Ti}^{Ia}, \theta_T^{non-Ia}, \theta_{Ti}^{non-Ia},
\theta_{gi})$.  The source model fixes
the luminosity. The  model includes  information on the
source and line-of-sight effects, which specify the effective SED.   Only the
parameters of the model of the appropriate type $T$ are relevant.  The galaxy parameters
$\theta_{gi}$ are included if correlations with host-galaxy properties are to be modeled.
\item $n_i(t,\lambda; L_i, \mu_i, z_i)$.  The  flux incident at Earth
is fixed by the luminosity, distance modulus, and redshift.
\end{itemize}

The model expectation of measured  signals is governed by:
\begin{itemize}
\item The transmission functions $\phi(\lambda)$ describe the optical path from the
atmosphere to counts.  We expect
the calibration group to provide us with the functions and covariance
$\phi \sim \mathcal{N}(\bar{\phi},C_\phi)$. Self-calibration, taking advantage of the
fact that SNe~Ia are excellent standard stars, 
may be included in this model.
\item $\mathit{ADU}$ and
$\mathit{ADU}_g$ are the expected counts of the transient and galaxy respectively,
coming from $\int n_i \phi d\lambda$.
\end{itemize}

The observations, denoted with \textasciicircum, are now discussed.
\begin{itemize}
\item $\hat{\mathit{ADU}}$ is the realized flux.  $\hat{\mathit{ADU}} \sim \mathcal{N}(ADU,C_{ADU})$
where $C_{ADU}$ is the photometric noise.
\item $\hat{T}_S$, $\hat{z}_S$, $\hat{\theta}_S$ Transient redshift and properties from
spectroscopic data. I skip the details of the actual measurement  and make a direct association
with the fluxes of the transient and underlying galaxy $n$, $n_g$.
\item $\hat{Z}$ represents the calibration measurements.  Although we expect to receive
the transmission functions from the Calibration Task Force, we may be want to include
calibration as part of our analysis.
\end{itemize}

@@

%The host-galaxy properties depend on the galaxy catalog and the transient
%coordinates $P(\hat{z}_H, \hat{\theta}_H | \hat{\mathit{Gals.}}, \text{RA/Dec})$.  With a

@@

%%$\left. \langle \hat{\mu}-\mu(z)\rangle \right|_{\text{mismatch}}$, $\left. \langle \hat{z}-z \rangle \right|_{\text{mismatch}}$ and cross-terms to correct bias in the Hubble diagram. 

@@

%P(\hat{D}_i=1,\hat{\mathit{ADU}}_{i}, \hat{T}_{Si}, \hat{z}_{Si}, \hat{\theta}_{Si}, \hat{z}_{Hi}, \hat{\theta}_{Hi}

@@

%The model contains a piece $P(\hat{\mathit{ADU}}_{i} \hat{z}_{Si}| \mu_i,z_i,\theta_T, \theta_{Ti}, ,\phi(\lambda))$, which

@@

%There term $P(\hat{D}|\mathit{ADU}_V)$.  For all high-redshift supernova searches

@@

%Generically $\vec{\hat{T}}$ can describe sample selection.  I choose to focus

>>>>

\documentclass[preprint]{aastex}
\usepackage{amsmath,amssymb}
\usepackage{mathrsfs}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{bm}
\bibliographystyle{apj}

@@

\title{Model for DES Type Ia Supernova Cosmology Analysis}
\author{Alex Kim}

\section{Model}
We infer a Hubble diagram based on measurements of discovered transients
classified as Type~Ia supernovae (SNe~Ia).  The probability distribution
function (pdf) of the Hubble diagram is denoted as
\begin{equation}
p({\mu},{z} |  {{ADU}}, {{T}}_S,{{z}}_S,
{{\theta}}_S,\text{GC}, \text{Detected}, {\text{Selected}},
{\text{RA}}, {\text{Dec}}).
\label{hd:eqn}
\end{equation}
The set of distance moduli
and redshifts $(\mu, z)$ is the Hubble diagram.  The available information
from which the Hubble diagram is inferred include:
the transient photometry ${ADU}$; spectral measurements of
transient
type, redshift and SN-parameters ${T}_S$, ${z}_S$, ${\theta}_S$;
the galaxy catalog ${\text{GC}}$; transient coordinates  $\text{RA}$, $\text{Dec}$;
and the fact that the transients were Detected and Selected for the analysis.

The pdf in Equation~\ref{hd:eqn} is proportional to
\begin{equation}
p({\mu},{z},   \text{Detected}, {\text{Selected}},
\text{RA}, \text{Dec} |  {{ADU}}, {{T}}_S,{{z}}_S,
{{\theta}}_S, \text{GC})
\label{hd2:eqn}
\end{equation}
and we model this scenario.  A sketch of such a model is depicted in the Probabilistic Graphical Model
(PGM)
shown in Figure~\ref{pgm:fig}.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=7in]{/Users/akim/project/abc/results/hdpgm.eps} 
   \caption{Probabilistic Graphical Model for the SN~Ia analysis.  
   The forward modeling
   flow goes from left to right, starting with the description of the universe, observatory,
   and data.
   A GC ${\mathit{Gal.}}$ informs the model for potential hosts $G$.  A transient
   at coordinates  $\text{RA/Dec}$  has a host $g_i$ drawn from $G$.
   The galaxy redshift redshift $z$ and parameters $\theta_{gi}$ depend on the host.
   A distance modulus model with parameters $\theta_\mu$ fix the distance modulus $\mu_i$.
   The transient type $T$ depends on the host-galaxy.   The transient
   parameters $\theta_T^X$, $\theta_{Ti}^X$ (and perhaps the host galaxy) fix the luminosity $L$.       The 
   incoming photon flux $n$, $n_g$  are then fixed
   with redshift and distance modulus.
   The instrumental transmission function $\phi$ is calibrated with data ${Z}$ and
   gives the expected
   counts $\overline{\mathit{ADU}}$, $\overline{\mathit{ADU}}_g$.  The galaxy catalog is GC. 
   The realized light curve (${ADU}$) depends on the expected counts and also
   must have been Detected and Selected.  Some transients have spectral data
   ${X}_S$.
   \label{pgm:fig}}
\end{figure}

The supernova survey provides the transients that are to be analyzed providing
the following information:
\begin{itemize}
\item ${\text{RA}}$, ${\text{Dec}}$.  The observed
coordinates.  For our purposes the observed, rather than true, coordinates are sufficient.
\item Detected.  The transient is discovered.
\item Selected. The transient is selected for the cosmology analysis.  This presumably
includes its classification as SN~Ia.
\item GC.  The galaxy catalog used to make host-galaxy associations.
\end{itemize}

The model describes the universe and its contents, specifically transients
that are discovered as typed as SN~Ia and their host galaxies.
\begin{itemize}
\item $G$ is a shorthand for the model of all galaxies that can host a transient.  It is
the union of the objects in the galaxy catalog GC
plus unobserved galaxies that have not
met the detection threshold, presumably due to their low surface brightness.
\item $P(g_i | G, {\text{RA}}/{\text{Dec}})$. The host galaxy $g_i$
is drawn from the GC based on coordinates.
This pdf is
the deliverable we want from Ravi.
\item $z_i$, $\theta_{gi}$ are the galaxy redshift and parameters (e.g.\ mass, sSFR, metallicity).
For observed galaxies the parameter values are based on measurements e.g.\ $\theta_{gi} \sim \mathcal{N}({\theta}_{gi},\sigma^2_{{\theta}_{gi}})$, otherwise the distribution
of $\theta_{gi}$ for ``hostless'' galaxies is described by the model.
Redshifts may be photometric or spectroscopic.
\item $\theta_\mu$ are the parameters that specify that distance modulus.
\item $\mu_i=\mu(z_i; \theta_\mu)$.  The distance modulus is a function of redshift,
we don't impose a specific form for  $\mu$ at the moment,  it could
be a $\Lambda$CDM prediction with cosmological parameters, a spline with knot values
as parameters, or as extreme as independent values for each $z$.
The redshift of the galaxy is used in the Hubble diagram.
\item $P(T_i | \theta_{gi})$.  The transient type $T_i$, SN~Ia or non-Ia, depends
on the host properties.  We could make the relative rates be dependent
on specific star formation rate and morphology.
\item $p(\theta_T^{Ia}, \theta_{Ti}^{Ia})$.  The model parameters for SNe~Ia, including
parameters that describe distributions.
Different SN~Ia models could be considered: the treatment of SALT2 parameters
has been presented in \citet{2011MNRAS.418.2308M}.
The SALT2 SN~Ia
model has individual $c$ and $x_1$ parameters that would constitute $\theta_{Ti}^{Ia}$,
and global $\alpha$, $\beta$, and intrinsic dispersion  parameters  common to all supernovae
 and  underlying  $c$ and $x_1$ distributions that would constitute $\theta_T^{Ia}$.
\item $p(\theta_T^{non-Ia}, \theta_{Ti}^{non-Ia})$.  Non-Ia transients
that can be mistyped as Ia are described by a parameterized model.
\item $L_i(t,\lambda; T_i, \theta_T^{Ia}, \theta_{Ti}^{Ia}, \theta_T^{non-Ia}, \theta_{Ti}^{non-Ia},
\theta_{gi})$.  The source model fixes
the luminosity. The  model includes  information on the
source and line-of-sight effects, which specify the effective SED.   Only the
parameters of the model of the appropriate type $T$ are relevant.  The galaxy parameters
$\theta_{gi}$ are included if correlations with host-galaxy properties are to be modeled.
\item $n_i(t,\lambda; L_i, \mu_i, z_i)$, $n_{gi}(t,\lambda; L_i, \mu_i, z_i)$.  The  flux incident at Earth
is fixed by the luminosity, distance modulus, and redshift.
\end{itemize}

The model expectation of measured  signals is governed by:
\begin{itemize}
\item The transmission functions $\phi(\lambda)$ describe the optical path from the
atmosphere to counts.  We expect
the calibration group to provide us with the functions and covariance
$\phi \sim \mathcal{N}(\bar{\phi},C_\phi)$. Self-calibration, taking advantage of the
fact that SNe~Ia are excellent standard stars, 
may be included in this model.
\item $\overline{\mathit{ADU}}$ and
$\overline{\mathit{ADU}}_g$ are the expected counts of the transient and galaxy respectively,
coming from $\int n_i \phi d\lambda$.
\end{itemize}

The observations, denoted with \textasciicircum, are now discussed.
\begin{itemize}
\item ${\mathit{ADU}}$ is the realized flux can be expressed as
\begin{equation}
P({\mathit{ADU}} | \text{Dectected}, \text{Selected}, \overline{\mathit{ADU}}, 
\overline{\mathit{ADU}}_g)=\frac{P({\mathit{ADU}},\text{Dectected}, \text{Selected} |  
\overline{\mathit{ADU}},  \overline{\mathit{ADU}}_g)}{P(\text{Dectected}, \text{Selected} | 
\overline{\mathit{ADU}},  \overline{{\mathit{ADU}}}_g)},
\end{equation}
which can be determined from detection and typing efficiency studies.
\item ${T}_S$, ${z}_S$, ${\theta}_S$ Transient type, redshift and properties from
spectroscopic data. I skip the details of the actual measurement  and make a direct association
with the fluxes of the transient and underlying galaxy $n$, $n_g$.
A very simple association is ${T}_S = T$, ${z}_S=z$, ${\theta}_S \in \theta_g$,
or a blurry version instead of a straight equality.
\item ${Z}$ represents the calibration measurements.  Although we expect to receive
the transmission functions from the Calibration Task Force, we may be want to include
calibration as part of our analysis.
\end{itemize}

@@

%The host-galaxy properties depend on the GC and the transient
%coordinates $P({z}_H, {\theta}_H | {\mathit{Gals.}}, \text{RA/Dec})$.  With a

@@

%%$\left. \langle {\mu}-\mu(z)\rangle \right|_{\text{mismatch}}$, $\left. \langle {z}-z \rangle \right|_{\text{mismatch}}$ and cross-terms to correct bias in the Hubble diagram. 

@@

%P({D}_i=1,{\mathit{ADU}}_{i}, {T}_{Si}, {z}_{Si}, {\theta}_{Si}, {z}_{Hi}, {\theta}_{Hi}

@@

%The model contains a piece $P({\mathit{ADU}}_{i} {z}_{Si}| \mu_i,z_i,\theta_T, \theta_{Ti}, ,\phi(\lambda))$, which

@@

%There term $P({D}|\mathit{ADU}_V)$.  For all high-redshift supernova searches

@@

%Generically ${{T}}$ can describe sample selection.  I choose to focus
